---
title: "Day 2: Intro to RNA-seq Supplementary Notebook"
author: Ally Hawkins
date: 2021-09-21
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

## Introduction

This notebook addresses some questions that came up after the second day of the CCDL workshop.

- [When to use quotations](#when-to-use-quotations)

### How to use this notebook: 

While it is fine to just read this page as is, if you want to get the Rmarkdown notebook itself to work with, you can follow the following steps:

1. Click the `Code` button in the upper right of this page and choose `Download Rmd` to save this notebook (`2021-09-21_muddiest_points.Rmd`) to your computer.
If you have RStudio installed on your computer, you can then open the file directly.

Otherwise you can add it to the CCDL RStudio server to work with it there with the following additional steps:

2. Navigate to RStudio server.  
3. In the `File` pane, navigate to `training-modules/RNA-seq`.     
4. Click the `Upload` button in the `File` pane.   
5. Click `Choose file` and find this `Rmd` file on your computer, then click `OK`  


## When to use quotations

Some questions came up in the slack about when to use and when not to use quotations in R. 
In general, the rule of thumb is that quotations are reserved for characters and strings (i.e. a word or set of words) only. 
When you put something in quotation marks, it's telling R to interpret the quotation marks as a string of characters.

Examples of places where you do NOT include quotations include: 

1. Assigning something to a variable - variable names should NOT be in quotation marks. 
2. When trying to access different columns of a dataframe if using a function from `tidyverse` such as `filter()` or `select()`, column names should NOT be in quotation marks. 

The following are a few examples of when you would use quotation marks: 

To assign a variable to a character or word, the variable will NOT be in quotes, but the string will be in quotes.

```{r}
# string is the variable being assigned 
string <- "this is a string"
```

When creating a vector of multiple strings, each element of the vector should be in quotes, but the variable that the vector is being assigned to should NOT be in quotes. 

```{r}
# each element in the vector is a string so should be in quotes 
animals_vector <- c("cat", "dog", "pig", "cow", "bird")
```

If you are checking to see if a vector contains a string, the string that you are searching for must be in quotes, but the vector is a variable and will not be in quotes. 

```{r}
"cat" %in% animals_vector
```

Let's turn our vector into a one column data frame and use tidyverse to show you when you would and wouldn't use quotes there. 

```{r}
animals_df <- data.frame(animals = animals_vector)
```

Now, if we are trying to filter our dataframe to only contain rows that have values equal to the string "cat", we are going to use `dplyr::filter()`. 
Here, we want to give `dplyr::filter()` the column that we are filtering on, in this case, animals, and then the value that we are looking for in that column, in this example, "cat". 

To do this, the column name should NOT be in quotes, but the value that we are trying to match should be in quotes. 
The reason that "cat" should be in quotes is because we are trying to filter for rows that contain the string "cat", so we are indicating that it is a string by putting the quotation marks around it. 

Let's test this out. 

```{r}
animals_df %>%
  dplyr::filter(animals == "cat")
```
Here, you should see only one row with the value "cat". 

What if we hadn't used quotes around "cat"? 

```{r}
animals_df %>%
  dplyr::filter(animals == cat)
```
You should see an error here and that's because it's searching for the variable cat, since we didn't put quotes around it to indicate that it was a string. 

If we wanted to do it this way, we would have to make sure we assign `cat` to the string we are searching for first, and then filter. 

```{r}
cat = "cat"

animals_df %>%
  dplyr::filter(animals == cat)
```
We see this works, but doesn't make much sense to do with just one string. 
Let's try to do our filtering again, but this time we want to find anything that matches either a "cat" or "dog". 

```{r}
# first we make a variable that is a vector containing the strings we are interested in 
pets_vector <- c("cat", "dog")

# then search for animals that are in that vector 
animals_df %>%
  dplyr::filter(animals %in% pets_vector)
```
Here, we see that by using the `%in%` operator we are searching for any values in the animals column that match any values in `pets_vector`. 
We used quotes in `pets_vector` to indicate that each of those values are strings, but if we had not done that, it would have assumed each of those were a variable and would have thrown an error since we did not assign dog to a variable. 

In general, a good rule of thumb is to reserve quotation marks for use only when referring to a string or character and not when referring to a variable.


## Session Info
```{r}
sessionInfo()
```
