---
title: "Day 1: Intro to R & tidyverse Supplementary Notebook"
author: Josh Shapiro
date: 2021-09-20
output:   
  html_notebook: 
    toc: true
    toc_float: true
---


## Introduction

This notebook addresses some questions that came up after the first day of the CCDL workshop.

- [Installing packages](#installing-packages)
- [Logical vectors](#logical-vectors)

### How to use this notebook: 

While it is fine to just read this page as is, if you want to get the Rmarkdown notebook itself to work with, you can follow the following steps:

1. Click the `Code` button in the upper right of this page and choose `Download Rmd` to save this notebook (`2021-09-20_muddiest_points.Rmd`) to your computer.
If you have RStudio installed on your computer, you can then open the file directly.

Otherwise you can add it to the CCDL RStudio server to work with it there with the following additional steps:

2. Navigate to RStudio server.  
3. In the `File` pane, navigate to `training-modules/intro-to-R-tidyverse`.     
4. Click the `Upload` button in the `File` pane.   
5. Click `Choose file` and find this `Rmd` file on your computer, then click `OK`  


Now you can open up this file and run it on the Server. 

## Installing packages

We do not explicitly cover package installation during this training, mostly for time reasons, but also because we want to make sure everything is there and ready to go!

But of course when you go to work on your own machine, some R packages may not be installed, so you may need to know how to install them!
Usually, this is pretty straightforward, and we will cover the most common situations here.

There are two main sources for R packages that we use, [CRAN](https://cran.r-project.org/) and [Bioconductor](https://bioconductor.org/).
Installation of packages from the two sources is slightly different.

### CRAN packages

Most general use packages (not biology-specific) are available through CRAN, the Comprehensive R Archive Network.
To install a package from CRAN, you use the `install.packages()` function, which will install the requested package(s) as well as any packages they might depend on.
So if I wanted to install (or update) the `tidyverse` package and all of its dependencies, I would use the following command:

```{r, eval = FALSE}
install.packages("tidyverse")
```

If I had more than one package to install, I might use a vector of the package names (each enclosed by quotes):
```{r eval = FALSE}
install.packages(c("magrittr", "dplyr", "palmerpenguins"))
```

The first time you use `install.packages()` you may be prompted (in the Console) to choose a "mirror". 
You will need to enter any responses in the console.
Using option `0`, the "cloud", is usually safe, but you can choose another location if you prefer.

If you are using RStudio, there is also an "Install" button in the `Packages` tab, which will prompt you to enter the packages you want to install, but this will only work with CRAN packages.

### Bioconductor packages

To install [Bioconductor](https://bioconductor.org/) packages, you need to use a specialized installation function from the `BiocManager` package.
So the first step is to install that package from CRAN:

```{r eval = FALSE}
install.packages("BiocManager")
```

Once that is installed, you can use `BiocManager::install()` to install any packages that are part of Bioconductor and their dependencies. 
(In fact you could use this always: it will find CRAN packages as well!)

```{r, eval = FALSE}
BiocManager::install("DESeq2")
```

`BiocManager::install()` tends to be a bit more chatty than the base R installation; be on the lookout for prompts that you need to respond to in the console.

## Logical vectors

I briefly talked about logical (Boolean) vectors during the first session, but they can definitely be confusing at times, so here I will go into a bit more depth. 

First off, what is a logical vector? A logical vector is one that contains only the values `TRUE` and `FALSE` (and potentially `NA`).
Notice the all caps: R requires this. 

So the simplest way to create a logical vector is just like another vector, using `c()` but only those values:

```{r}
my_bool <- c(TRUE, FALSE, FALSE, TRUE)
my_bool
class(my_bool)
```
(Full disclosure, you _can_ use `T` and `F` in addition to `TRUE` and `FALSE`, but this is a bad idea, because it is possible to overwrite those abbreviations like normal variable names. `F <- TRUE` is a perfectly valid statement, even though it can wreak havoc on sloppy code.)


That isn't too useful, generally, because we usually want to create our logical vectors based on some data. To that we will tend to use one of a few comparison or logical operators:

Starting wiht the comparisons:
* `==`: equal to
* `<`: less than
* `<=`:  less than or equal to 
* `>`: greater than
* `>=`: greater than or equal to
* `%in%` : is a member of some other vector

Let's say we have a vector of (semi) random numbers. 

```{r}
rand_nums <- c(0.74, -0.83, 0.42, 0.29, -0.825, 0.25, 1.68)
```

We can use comparisons to create new logical vectors:

```{r}
negative <- rand_nums < 0
negative
```

Or we can test which of some character vectors match a particular value:
```{r}
words <- c("hello", "goodbye", "howdy", "aloha", "cat")
is_cat <- words == "cat"
is_cat
```

or are in a set of known values: 
```{r}
greetings <- c("aloha", "hello", "hi", "hola", "howdy")
is_greeting <- words %in% greetings
is_greeting
```

We can also modify combine multiple boolean vectors with logical statements to get new boolean vectors.

* `!`: not (inverts `TRUE` and `FALSE` )
* `|` : logical **or** - if either of the elements are `TRUE`, this gives `TRUE`
* `&` : logical **and** - if both of the elements are `TRUE`, this gives `TRUE`

These logical operators will act on each of the elements of a vector, or each pair from two vectors (note that I am not saving these to new vectors, but you could!):

```{r}
!is_greeting
```

```{r}
# Is it a greeting or is it a cat, because either way, we should say hi back
is_greeting | is_cat
```


Once we have a logical vector with the values we are after, we can use them in selections, much like we might use the position numbers, to get only those values that gave us `TRUE` in whatever comparison we were doing:

```{r}
greet_words <- words[is_greeting]
greet_words
```

We can also use them in many other places, such as the `dplyr::filter()` statements we covered in lecture.



## Session Info
```{r}
sessionInfo()
```





